HO unification from object language to meta language

### 2024-02-21 paper/main.tex
update tex

In order to reason about unification of the terms of an objet language within a
meta language, we start by formally describing the two languages. Employing
meta-programming for this purpose, \cref{code:lang-descr} presents, on the left,
the typing structure of object language terms along with the signature of the
\eqfo function.

@@

\subsection[Compilation: fo\_tm to tm]{Compilation: \elpiIn{fo_tm} $\to$ \elpiIn{tm}}

- how we transform an fo\_tm in tm
- the role of links

\subsection{Unification in ML}

- we accept HO unif with PF

\subsection{Term de-compilation}
% Compile FO to HO

% Assignement in HO should be decompiled

-------------------------------------------------------------------

\noindent

\subsection{implementation}

l'HO encoding e' esattamente lambda Prolog/elpi, il compilatore
in pratica potrebbe essere scritto in un meta language, qui lo si presenta in
elpi stesso. HO e FO in questo paper sono deep embedded in elpi per parlarne,
ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.

\section{recovering eta}

\begin{elpicode}
q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
F = fun a => app [f,a] ----> F = f
\end{elpicode}

l'utene da p su f, mentre l'istanza pe q forza F a fun ..

\section{recovering beta}

\begin{elpicode}
  q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
  F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
\end{elpicode}

qui la sintesi di F puo generare un beta redex, quindi ci mettiamo
p1 F1, e decomp beta F [a] F1.

\section{recovering eta-beta within unification (non linear variables)}

se i problemi di cui sopra avvengono nello stesso termine

\begin{elpicode}
  q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
\end{elpicode}

bisogna slegare le due F e poi unificare le soluzioni tra di loro

\section{heuristic / binary app}

fo approx / sub pattern fragment

\begin{elpicode}
  p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
  p (all x\G x) F' = 
  G = x\ f x x a
  F = lam x\f x x
  F' = (app[f,b,b])
  link (F a) F'
  link G F
\end{elpicode}

% questo esempio recupera l'app binaria perche

\begin{elpicode}
  (app (app F x) a) = (app (f x x) a)
\end{elpicode}

\printbibliography

>>>>

In order to reason about unification of the terms of the objet language within the
meta language, we start by formally describing the two languages. Employing
meta-programming for this purpose, \cref{code:lang-descr} presents the type
\elpiIn{tm} containing the constructors for term application, lambda abstraction
and constants. Moreover, in order to represent unification variables, we need 
to give two different constructors.

In the case of the OL, variables have no scope. For example, the subterm
\coqIn{P x} from the instance forall\_dec. Here, \coqIn{P} is an higher order
variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} is a name bound to
\coqIn{P}. However, at the meta level, the translation of \coqIn{P x} becomes
\elpiIn{app[P, x]}, that is, \elpiIn{P} cannot reference \elpiIn{x}. In our
encoding, unification variables are encoded as integer corresponding to memory
addresses, and the constructor for unification variables in the OL is
\elpiIn{fo_uv}. This constructor is prefixed with \elpiIn{fo} since the
variable, having no scope, is a first-order variable.

On the other hand, in the case of the ML, we want to use its unification
algorithm to make variable assignment. Since the ML is an Higher Order
Programming Language, we represent unification variables with the \elpiIn{uv}
constructor, which, this time, can see a list of terms. Of course some attention
should be payed when dealing with this constructor, since we have to certify
each time that an \elpiIn{uv i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is a list of distinct names. Finally,
the following code

\begin{elpicode}
  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
\end{elpicode}

\noindent
illustrates the \elpiIn{assmt} representing variable assignment in the ML.

Variable assignment

\vspace{30pt}

@@

% \section[Compilation: fo\_tm to tm]{Compilation: \elpiIn{fo_tm} $\to$ \elpiIn{tm}}

% - how we transform an fo\_tm in tm \\
% - the role of links \\
% - decomp
% - esempio che va in questa semplice rappresentazione (from intro)
% - esempio che non va, multi-var, eta, beta
% % \section{Term de-compilation}

% \section{Unification in ML}

% - we accept HO unif with PF \\
% - need of multiple vars for a single OL var

% % Compile FO to HO

% % Assignement in HO should be decompiled

% -------------------------------------------------------------------

% \noindent

% \subsection{implementation}

% l'HO encoding e' esattamente lambda Prolog/elpi, il compilatore
% in pratica potrebbe essere scritto in un meta language, qui lo si presenta in
% elpi stesso. HO e FO in questo paper sono deep embedded in elpi per parlarne,
% ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.

% \section{recovering eta}

% \begin{elpicode}
% q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
% F = fun a => app [f,a] ----> F = f
% \end{elpicode}

% l'utene da p su f, mentre l'istanza pe q forza F a fun ..

% \section{recovering beta}

% \begin{elpicode}
%   q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
%   F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
% \end{elpicode}

% qui la sintesi di F puo generare un beta redex, quindi ci mettiamo
% p1 F1, e decomp beta F [a] F1.

% \section{recovering eta-beta within unification (non linear variables)}

% se i problemi di cui sopra avvengono nello stesso termine

% \begin{elpicode}
%   q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
% \end{elpicode}

% bisogna slegare le due F e poi unificare le soluzioni tra di loro

% \section{heuristic / binary app}

% fo approx / sub pattern fragment

% \begin{elpicode}
%   p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
%   p (all x\G x) F' = 
%   G = x\ f x x a
%   F = lam x\f x x
%   F' = (app[f,b,b])
%   link (F a) F'
%   link G F
% \end{elpicode}

% % questo esempio recupera l'app binaria perche

% \begin{elpicode}
%   (app (app F x) a) = (app (f x x) a)
% \end{elpicode}

% \section{Stats}

% Concretamente:

% - quante volte l'ordine superiore è necessario in stdpp/tlc

% \section{Conclusion}

% \printbibliography
